A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)

### 2014-03-14 paper-plos.tex
final

\Fixme{move this section back to the its original position if you don't like it
here (it was just before Acknowledgements)}

@@

\Change{%
The equivalent Hadoop-based pipeline was implemented with Crossbow. The input
data and the indexed genome reference were copied to Hadoop's storage
system~(HDFS) before starting the experiments.  Specifically, the storage volume
containing the input data was mounted with SSHFS to one of the Hadoop nodes,
from where the \texttt{bzip2}-compressed data were transferred to the HDFS\@.
This preparation was done to eliminate unpredictable effects on the experiments
from possible network delays.
}
Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program). Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from GitHub~\cite{code_repo_mr}). Our implementation is described in Section~\ref{sec:preprocessor}.

@@

\Change{%
All data was standardized to FASTQ format and compressed with \texttt{bzip2},
since it is typical to store sequencing data in a compressed file format. The
\texttt{bzip2} compression codec was selected it provides very good compression
and is splittable by Hadoop, meaning that the archive can be expanded in a
parallel manner allowing Hadoop to process multiple segments simultaneously.
Also, since some of the datasets lack a standard FASTQ record id, which is
required by our test pipelines to extract the read number, we also standardised
the read ids in the datasets by generating new ids based on the SHA-1 hash
function and labelled read mates with ``.1" and ``.2" suffixes.
}

@@

\Change{%
The Crossbow package includes a native preprocessing stage that provides great
flexibility in delivering the data to the Hadoop cluster; it can autonomously
download data from Amazon S3, FTP and HTTP servers over the
Internet~\cite{Langmead:2009kx}. However, this preprocessor does not parallelize
its operations. In theory one can manually introduce some degree of
parallelization by manually splitting the original input read data into smaller
files and to explicitly configuring Crossbow's preprocessor to download each one
(this is done by listing all of them in a special \textit{manifest file}).
Nevertheless, in practice this approach is rather error-prone and laborious.

%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
We reimplemented the main functionality in the Crossbow preprocessor as an MR
program to benefit from its massively parallel nature. The new MR preprocessor
was implemented in Python using the Hadoop streaming library.
\Fixme{Consider deleting the following implementation detail.}
To route both reads of the same pair to the same Reducer, a secondary
key-sorting mechanism was employed.
The resulting program efficiently processes short reads produced by different
versions of the Illumina sequencing platform as well as FASTQ files generated
from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
Hadoop-based approach imposes the requirement that the input data be stored
where the Hadoop nodes can access it, which is in our opinion a reasonable since
sequencing platforms generally have direct access to extensive storage and
computing facilities.
}

@@

\Change{%
As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
with a batch processing system (hereafter referred to as the HPC method)
(Methods).  Our experiments consisted of executing the pipelines for each of the
five input 
datasets (see Table~\ref{table:datasets}) and
measuring the wall-clock run time for each pipeline stage.  All experiments were
repeated
\Fixme{several times (how many?)}
and the results averaged to obtain a data point for the
particular combination of data size and computational platform.
}

\Fixme{Belongs with Discussion:
In light of our observations, we
propose modifications for existing Hadoop software to better exploit the
massively parallel nature of MR computations. Finally, we developed a set of Bash
scripts to parallelize the operations of classical DNA-seq analysis programs for
short-read alignment over multiple nodes and fully exploit the network bandwidth, thereby
accelerating calculations.
}

\subsection*{Accuracy of Pipelines}
\Change{%
Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
SOAPsnp~\cite{soapsnp}, respectively) we could not expect them to deliver perfectly
matching SNP lists. Instead, we verified the accuracy of the test pipelines by
verifying their ability to correctly identify the mutation $C\rightarrow T$ on
chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.
}

@@

\Change{%
To quantify the processing time saved by adopting the MR approach for
preprocessing the read data, we ran an
experiment comparing the new preprocessor described in
Section~\ref{sec:preprocessor}  to the original preprocessor furnished with
Crossbow.  To further validate our strategy, we also compared the performance of
our MapReduce approach to a different parallelization strategy based on a custom Bash script
that uses of all processing cores available on a single computing node to accelerate the operation (see
Supplementary Material for details of the implementation). This program read its
input directly from conventional HPC storage, rather than HDFS\@.  The experiment
was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
with 56 cores (7 slave nodes) while the Bash approach ran on a single node
equipped with 16 CPU cores.
}

@@

\Fixme{Alexey, please verify whether my interpretation seems reasonable. LP}

\Change{%
Table~\ref{table:preprocess} shows the results of the experiment. Clearly,
either of the parallel approaches are much faster than the original tool, with
all dataset sizes.  Also, we noticed that the difference between the Bash and Hadoop
approaches is minimal when running on Dataset II\@.  Indeed, with the relatively
small dataset size the overhead incurred by the Hadoop framework likely
overwhelms the useful compute time used for each available CPU core. However,
with increasing input size the superior scalability of MR becomes apparent, with
the largest dataset taking requiring less than half the time of the Bash
approach to be processed; at the same time, it is important to also notice that
the Hadoop approach employed approx.\ 1.6 times more CPU time compared to the
Bash approach.
}

@@

\subsection*{Scalability of HPC and Hadoop approaches}
\Change{%
Given the wide range in genome dataset sizes processed in practice -- e.g.,
exomes,targeted sequencing, whole-genome sequencing at various depths -- it is
important to understand the scaling characteristics of each processing
platforms (Hadoop and HPC) with respect to input size. For each platform, we
measured the time required for to execute their respective variant calling
pipeline (see Methods) for each of the test datasets (I-V). The results are
shown in Table~\ref{table:4}.

The measurements highlight that one of the advantages of using the Hadoop~MR} framework is
its almost linear scalability with respect to input size -- i.e.\ the
calculation time linearly depends on the size of the
dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
%Deleting this since one could dream/find pathological cases that don't work: ``
\Fixme{\sout{regardless the scaling nature of the underlying program.}}
Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the wall clock
running time as a function of the dataset size for the 56-core Hadoop cluster and datasets
I-IV.
%To stress the linear nature of scaling, both sets of points were fit using the
%least squares method.

\Change{%
Analogously, we investigated the scalability characteristics of the HPC method
with respect to input size.
\Fixme{The following works better the aid of a plot.}
Analyzing the observations in Table~\ref{table:4}, we can see with the HPC
method run time increases superlinearly as a function of dataset size.  As
expected, the result is that the HPC method is not able to complete the analysis
with our larger tests datasets.  On the other hand, this approach is not much
slower than our Hadoop cluster considering the much smaller quantity of computing
resources that is uses -- an issue that is analyzed more closely in the section
on the efficiency of the two systems.
}

\Fixme{The following paragraph comments on how the HPC approaches' runtime
varies with the number of cores available, rather than input size. If we're
going to leave it in, I think we should make an analogous discussion regarding
how Hadoop's runtime varies with increasing resources. The data in Table 6 might
be suitable for this.}

@@

\subsection*{Comparing the efficiency of Hadoop and HPC}
\Change{%
Hadoop was designed to process huge datasets by distributing computations to a
large number of computers\cite{hadoop,lin2010}.  Therefore, while allowing us to
process very large datasets, the mechanisms required for distributing the work
add a non-trivial overhead to the computation, which becomes more and more
significant as the amount of work to be done is reduced -- i.e.\ the dataset
size decreases.

In order to measure the overhead incurred by the Hadoop platform, we compared
the efficiency of the two processing methods (Hadoop and HPC) as the size of the
dataset varies. We measure the efficiency indirectly as the number of core-hours
($F$) used for same analysis: $F=T_{p}\times p,$ where~$T_{p}$ is the running time on~$p$ cores.
}

\Change{%
\Fixme{The way of reorganized this text, at the end of the subsection it kind of
excessively teeter-totters too much between HPC and Hadoop.}

Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the
ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
%
The simplest curve that fits the data points nicely is a parabolic function
which, importantly, is not a constant.  Since our measurements using the
Hadoop-based pipeline, summarized in Figure~\ref{fig:fig1}, showed linear
scaling properties of Hadoop in this setting, a deviation of the ratio
$F_{Hadoop}/F_{HPC}$ from the constant is evidence of the lack of scalability of
the HPC approach.
}%
%Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e. being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}. \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?} 
We suspect that the main reason for such a behavior is that for large datasets
(Dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory-
and time-demanding than the alignment with Bowtie.  Since its memory
requirements scale with input size, with sufficient input data Samtools swaps
data to disk, creating multiple (up to hundreds) temporary files while it sorts
the BAM file -- incidentally, these files can end up on shared network-accessed
file systems and thus generate large amounts of network traffic.
Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with
different amount of RAM available on a single HPC node. Though there is a large
deviation when using small amounts of RAM -- we believe due to other users' load
on the cluster network -- the trend quite clearly indicates that a relatively
large amount of RAM per node is needed in order to use Samtools effectively. In
contrast, in our tests Hadoop performed the SNP calling on much more economic
hardware, in linear time, with just 12GB of RAM per \textit{virtual} node.

\Change{%
Nevertheless, it is evident from the plot that the HPC method is always more
efficient in its use of CPU time than the Hadoop-based approach, and thus takes fewer computing
resources to perform the same amount of work.  However, Hadoop becomes
increasingly more efficient with respect to the HPC method as the dataset size
increases.
}
However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
\Change{%
In addition it is quite important to note that the price paid for
the efficiency in HPC's method is in overall runtime and in its inability to
finish the analysis on Dataset~V in a reasonable time (over 48 hours to Hadoop's 19.4
hours; see Table~\ref{table:4}).
}

@@

The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages. The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
To study this phenomenon, we compared HPC and Hadoop network communication costs
\Fixme{(maybe ``network load''?)}
as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).

@@

There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
%
However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
\Fixme{(so, after reading the previous sentence, the logical question is ``why
not use gnumap?'')}
%
To circumvent this limitation we implemented a highly optimized Bash script
based on standard Unix utilities that makes very efficient use of the HPC
cluster network~\cite{code_repo_bash}; we then compared the network
performance with the standard Hadoop HDFS approach. We separated alignment
time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in
FASTQ format have to be preprocessed (involving node communication~$T_{comm}$)
to be able to run in MR-fashion, while the data locality will be automatically
achieved during the data ingestion into the HDFS\@.  We rewrote the code for the
preprocessing stage for Crossbow to make it suitable for MR-style
parallelization.  For the HPC approach, $T_{comm}$ involves the chunks that are
sent from the sequence delivery location to the local node scratch disks where
the actual alignment happens, and the sending of the aligned SAM files back to
the delivery location over the network.
%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}

@@

%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@. One can see two distinct linear stretches. One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60. The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed. The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.} 
and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot. 
The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec. Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch. The nodes are being allocated on the same rack, which is the default SLURM behavior.

@@

\Fixme{Galaxy seems a bit out of place to me.
\sout{A popular way to define and run bioinformatics pipelines on HPC resources
is via the} Galaxy\cite{galaxy} \sout{platform, which provides a Web-based graphical
user interface~(GUI) to bioinformatic programs, simplifying the experience for
the end user.}
}

@@

In this report we have described two approaches for high-performance analysis of
DNA sequencing data; one based on regular HPC using a batch system and one based
on Hadoop. We show that Hadoop installed on a private cloud is an appealing
solution both on terms of runtime performance and also in terms of usability. A
key result is the data size where Hadoop becomes an advantageous option as compared
to regular HPC batch systems.
In order to compare the two platforms we developed highly optimized pipelines
for both scenarios. We found that Hadoop excels over HPC when processing
datasets larger than 100\,Gbases. Also, Hadoop shows almost linear scalability for
this problem. Extrapolation to infinite dataset size reveals however that HPC
provides the results faster, given the same amount of resources as Hadoop on a
private cloud. \Fixme{But the HPC method tested ran on a single node so it
didn't incur distribution costs!}

We implemented a new MapReduce program to perform the preprocessing stage
required by Crossbow in order to enable to Hadoop-based pipeline to effectively
use the available computing resources. We also wrote a highly optimized Bash
script to compare the scaling relations between the ratio of the alignment time
to the communication time as a function of the reciprocal number of nodes on an
HPC system.  Our results show that the calculations on Hadoop with HDFS scales
better than the network attached parallel storage commonly used in the HPC
centers.  In addition, we showed how performance of the HPC approach can be
improved by redefining the queueing system's default behavior. This is however
not of practical use in an HPC center since the network bandwidth is shared
among multiple users. Finally, we demonstrated that an existing and extensible
publicly available web-based GUI (Cloudgene) provides an easy way to execute
bioinformatics analysis on Hadoop for those who are less experienced with Linux
scripting.

@@

\begin{verbatim}
pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}
\end{verbatim}

@@

\begin{figure}[!ht]
	\input{./figures/fig1/fig1.tex}
	\caption{%
	\Change{%
		Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
		increasing input dataset size. Each test dataset (labelled with Roman
		numerals) was used. The image clearly highlights the linear scaling
		characteristics of the Hadoop-based approach.
		}
		\Fixme{\\ What would you think of inserting line plotting the 16-core HPC
		tests from Table 4}
		}
	\label{fig:fig1}
\end{figure}

\begin{figure}[!ht]
	\input{./figures/fig2/fig2.tex}
	\caption{%
	\Change{%
	The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
	a 16-core HPC node, respectively.
	The points are fit to a quadratic least-squares curve, which makes it possible
	to predict {\it infinite} dataset size. Datasets are labeled with Roman
	numerals.
	}
	}
	\label{fig:fig2}
\end{figure}

@@

\caption{%
	\Change{%
	Time (in minutes) required by Samtools to sort
	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
	amount of available RAM reported is for all 16 cores.
	}
}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
RAM, GB		&	8			&		16			&			32		& 		64				\\
timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
\end{tabular}
\end{center}
\label{table:5}
\normalsize
\end{table}%

\begin{table}[!ht]
\small
\caption{%
	\Change{%
	Time (in minutes) employed to preprocess Datasets II, III, and IV by
	Crossbow's native read preprocessor and our MR preprocessor on a 56-core Hadoop
	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
	}
}

>>>>

@@

The equivalent Hadoop-based pipeline was implemented with Crossbow. The input
data and the indexed genome reference were copied to Hadoop's storage
system~(HDFS) before starting the experiments.  Specifically, the storage volume
containing the input data was mounted with SSHFS to one of the Hadoop nodes,
from where the \texttt{bzip2}-compressed data were transferred to the HDFS.
Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program). Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from GitHub~\cite{code_repo_mr}). Our implementation is described in Section~{\it MapReduce-based Data Preprocessor}.

@@

All data was standardized to FASTQ format and compressed. The
\texttt{bzip2} compression codec was selected as it provides very good compression
and is splittable by Hadoop, meaning that the archive can be expanded in a
parallel manner allowing Hadoop to process multiple segments simultaneously.
Also, due to  lack a standard FASTQ record id, which is
required by our test pipelines to extract the read number, we also standardised
the read ids in the datasets by generating new ids based on the SHA-1 hash
function and labelled read mates with ``.1" and ``.2" suffixes.

@@

The Crossbow package includes a native preprocessing stage that provides great
flexibility in delivering the data to the Hadoop cluster; it can autonomously
download data from Amazon S3, FTP and HTTP servers over the
Internet~\cite{Langmead:2009kx}. However, this preprocessor does not parallelize
its operations. In theory one can manually introduce some degree of
parallelization by manually splitting the original input read data into smaller
files and to explicitly configuring Crossbow's preprocessor to download each one
(this is done by listing all of them in a special \textit{manifest file}).

%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
We reimplemented the main functionality in the Crossbow preprocessor as an MR
program to benefit from its massively parallel nature. The new MR preprocessor
was implemented in Python using the Hadoop streaming library.
%\Fixme{Consider deleting the following implementation detail.}
%To route both reads of the same pair to the same Reducer, a secondary key-sorting mechanism was employed.
The resulting program efficiently processes short reads produced by different
versions of the Illumina sequencing platform as well as FASTQ files generated
from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
Hadoop-based approach imposes the requirement that the input data be stored
where the Hadoop nodes can access it, which is in our opinion a reasonable since
sequencing platforms generally have direct access to extensive storage and
computing facilities.

@@

As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
with a batch processing system (hereafter referred to as the HPC method)
(Methods).  Our experiments consisted of executing the pipelines for each of the
five input 
datasets (see Table~\ref{table:datasets}) and
measuring the wall-clock run time for each pipeline stage.  All experiments were
repeated at least three times, 
and the results averaged to obtain a data point for the
particular combination of data size and computational platform.

\subsection*{Accuracy of Pipelines}
Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
SOAPsnp~\cite{soapsnp}, respectively) we could not expect them to deliver perfectly
matching SNP lists. Instead, we verified the accuracy of the test pipelines by
verifying their ability to correctly identify the mutation $C\rightarrow T$ on
chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.

@@

To quantify the processing time saved by adopting the MR approach for
preprocessing the read data, we ran an
experiment comparing the new preprocessor described in
Section~{\it MapReduce-based Data Preprocessor} to the original preprocessor furnished with
Crossbow.  To further validate our strategy, we also compared the performance of
our MapReduce approach to a different parallelization strategy based on a custom Bash script
that uses of all processing cores available on a single computing node to accelerate the operation (see
Supplementary Material for details of the implementation). This program read its
input directly from conventional HPC storage, rather than HDFS\@.  The experiment
was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
with 56 cores (7 slave nodes) while the Bash approach ran on a single node
equipped with 16 CPU cores.

@@

Table~\ref{table:preprocess} shows the results of the experiment. Clearly,
either of the parallel approaches are much faster than the original tool, with
all dataset sizes.  Also, we noticed that the difference between the Bash and Hadoop
approaches is minimal when running on Dataset II\@.  Indeed, with the relatively
small dataset size the overhead incurred by the Hadoop framework likely
overwhelms the useful compute time used for each available CPU core. However,
with increasing input size the superior scalability of MR becomes apparent, with
the largest dataset taking requiring less than half the time of the Bash
approach to be processed; at the same time, it is important to also notice that
the Hadoop approach employed approx.\ 1.6 times more CPU time compared to the
Bash approach.

@@

\subsection*{Scalability of HPC and Hadoop approaches}
Given the wide range in genome dataset sizes processed in practice -- e.g.,
exomes, targeted sequencing, whole-genome sequencing at various depths -- it is
important to understand the scaling characteristics of each processing
platforms (Hadoop and HPC) with respect to input size. For each platform, we
measured the time required for to execute their respective variant calling
pipeline (see Methods) for each of the test datasets (I-V). The results are
shown in Table~\ref{table:4}.

The measurements highlight that one of the advantages of using the Hadoop~MR framework is
its almost linear scalability with respect to input size -- i.e.\ the
calculation time linearly depends on the size of the
dataset, regardless the scaling nature of the underlying program~\cite{Langmead:2009kx,Pireddu:2011vn}.
Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the wall clock
running time as a function of the dataset size for the 56-core Hadoop cluster and datasets
I-IV.
%To stress the linear nature of scaling, both sets of points were fit using the
%least squares method.

@@

\subsection*{Comparing the efficiency of Hadoop and HPC}
Hadoop was designed to digest huge datasets\cite{hadoop,lin2010}. One can propose then that the larger the dataset is, the more suitable Hadoop becomes compared to the HPC approach. In order to compare the ``suitability'' for different types of calculation platforms (Hadoop and HPC) each being run on a different number of cores, we constructed the function showing the amount of used core-hours:
$F=T_{p}\times p,$ 
where~$T_{p}$ is the calculation time on~$p$ cores.

Using the data from Table~\ref{table:4}, we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size (Figure~\ref{fig:fig2}). Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset. As one can see, the Hadoop approach becomes more and more effective compared to the HPC scenario as the dataset size increases. 
The parabolic extrapolation for our settings give the ratio as~$1.70\pm0.01$, meaning that Hadoop running even in the virtualized environment of a private cloud assembled on moderate hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets greater than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.

The simplest curve to fit the points nicely is a parabolic function, importantly not a constant. 
Since the Hadoop approach for the datasets reveals the linear scaling (Figure~\ref{fig:fig1}), a deviation of the ratio from the constant shows the 
lack of scalability of the HPC approach.
%Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e. being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}. \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?} 
We suspect that the main reason for such a behavior is that for large datasets
(Dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory-
and time-demanding than the alignment with Bowtie.  Since its memory
requirements scale with input size, with sufficient input data Samtools swaps
data to disk, creating multiple (up to hundreds) temporary files while it sorts
the BAM file -- incidentally, these files can end up on shared network-accessed
file systems and thus generate large amounts of network traffic.
Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with
different amount of RAM available on a single HPC node. Though there is a large
deviation when using small amounts of RAM -- we believe due to other users' load
on the cluster network -- the trend quite clearly indicates that a relatively
large amount of RAM per node is needed in order to use Samtools effectively. In
contrast, in our tests Hadoop performed the SNP calling on much more economic
hardware, in linear time, with just 12GB of RAM per \textit{virtual} node.

Nevertheless, it is evident from the plot that the HPC method is always more
efficient in its use of CPU time than the Hadoop-based approach, and thus takes fewer computing
resources to perform the same amount of work.  However, Hadoop becomes
increasingly more efficient with respect to the HPC method as the dataset size
increases.

However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
In addition it is quite important to note that the price paid for
the efficiency in HPC's method is in overall runtime and in its inability to
finish the analysis on Dataset~V in a reasonable time (over 48 hours to Hadoop's 19.4
hours; see Table~\ref{table:4}).

@@

The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages. The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
To study this phenomenon, we compared HPC and Hadoop network communication costs
as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).

@@

There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
%
However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.%
To circumvent this limitation we implemented a highly optimized Bash script
based on standard Unix utilities that makes very efficient use of the HPC
cluster network~\cite{code_repo_bash}; we then compared the network
performance with the standard Hadoop HDFS approach. We separated alignment
time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in
FASTQ format have to be preprocessed (referred to as~$T_{comm}$)
to be able to run in MR-fashion, while the data locality will be automatically
achieved during the data ingestion into the HDFS\@.  We rewrote the code for the
preprocessing stage for Crossbow to make it suitable for MR-style
parallelization.  For the HPC approach,~$T_{comm}$ involves the chunks that are
sent from the sequence delivery location to the local node scratch disks where
the actual alignment happens, and the sending of the aligned SAM files back to
the delivery location over the network.
%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}

@@

%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@. One can see two distinct linear stretches. One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60. The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed. The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.} 
and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency in the selected coordinates, which can be observed in the plot. 
The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec. Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch. The nodes are being allocated on the same rack, which is the default SLURM behavior.

@@

A popular way to define and run bioinformatics pipelines on HPC resources
is via the Galaxy\cite{galaxy} platform, which provides a Web-based graphical
user interface~(GUI) to bioinformatic programs, simplifying the experience for
the end user.

@@

In this report we have described two approaches for high-performance analysis of
DNA sequencing data; one based on regular HPC using a batch system and one based
on Hadoop. We show that Hadoop installed on a private cloud is an appealing
solution both on terms of runtime performance and also in terms of usability. A
key result is the data size where Hadoop becomes an advantageous option as compared
to regular HPC batch systems.
In order to compare the two platforms we developed highly optimized pipelines
for both scenarios. We found that Hadoop becomes comparable with HPC when processing
datasets larger than 100\,Gbases. Also, Hadoop shows almost linear scalability for
this problem. Extrapolation to infinite dataset size reveals however that HPC
provides the results faster, given the same amount of resources as Hadoop on a
private cloud.

We implemented a new MapReduce program to perform the preprocessing stage
required by Crossbow in order to enable to Hadoop-based pipeline to effectively
use the available computing resources. We also wrote a highly optimized Bash
script to compare the scaling relations between the ratio of the alignment time
to the communication time as a function of the reciprocal number of nodes on an
HPC system.  Our results show that the calculations on Hadoop with HDFS scales
better than the network attached parallel storage commonly used in the HPC
centers.  In addition, we showed how performance of the HPC approach can be
improved by redefining the queueing system's default behavior. Finally, we demonstrated that an existing and extensible
publicly available web-based GUI (Cloudgene) provides an easy way to execute
bioinformatics analysis on Hadoop for those who are less experienced with Linux
scripting.

@@

\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc \${file2}) 
| paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}

@@

\begin{figure}[!ht]
	\input{./figures/fig1/fig1.tex}
	\caption{%
		Runtime for the Hadoop-based pipeline on the 56-core Hadoop cluster with
		increasing input dataset size. Each test dataset (labelled with Roman
		numerals) was used. The image clearly highlights the linear scaling
		characteristics of the Hadoop-based approach.
		}
	\label{fig:fig1}
\end{figure}

\begin{figure}[!ht]
	\input{./figures/fig2/fig2.tex}
	\caption{%
	The ratio of the~$F_{Hadoop}/F_{HPC}$ as a function of a reciprocal
	dataset size in Gbases. The pipelines were run on a 56-core Hadoop cluster and
	a 16-core HPC node, respectively.
	The points are fit to a quadratic least-squares curve, which makes it possible
	to predict {\it infinite} dataset size. Datasets are labeled with Roman
	numerals.
	}
	\label{fig:fig2}
\end{figure}

@@

\caption{%
	Time (in minutes) required by Samtools to sort
	Dataset IV (BAM file containing 58 Gbases). The measurements are done on a single HPC
	node with $p=16$ cores. Averaging is done over 5 independent simulations. The
	amount of available RAM reported is for all 16 cores.
}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
RAM, GB		&	8			&		16			&			32		& 		64				\\
timing, minutes	&	$208\pm50$	&	$172\pm13$		&	$136\pm12$		& 	$87\pm11$			\\
\end{tabular}
\end{center}
\label{table:5}
\normalsize
\end{table}%

\begin{table}[!ht]
\small
\caption{%
	Time (in minutes) employed to preprocess Datasets II, III, and IV by
	Crossbow's native read preprocessor and our MR preprocessor on a 56-core Hadoop
	cluster, and by the custom multi-core Bash script on a single 16-core HPC node.
	}
